Simulating High-Dimensional Multivariate Data

### 2021-01-13 010-background.Rmd
2nd draft. Polishing ch050 after new RNA-seq

```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
if ( !file.exists( './data/exampleGenes.rds' ) )  {
    set.seed(01112021)
    exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
    saveRDS(exampleGenes, file = './data/exampleGenes.rds')
}
exampleGenes <- readRDS('./data/exampleGenes.rds')
## Add row names per WWP comment
small <-  brca %>% select( exampleGenes ) %>% as.data.frame
numPatients <- 5
smallForTable <- head( small, n = numPatients )
smallForTable$PatientID <- head( readLines('./data/examplePatients.txt'), n = numPatients )
smallForTable <- smallForTable[ , c(4, 1:3 ) ]
smallForTable[ , 2:4 ] <- round( smallForTable[ , 2:4 ] )
knitr::kable(smallForTable ,
      booktabs = TRUE,
      caption = "mRNA expression for three selected high-expressing genes (*RPL5*, *TKNIP*, and *VIM*) for the first five patients in the TCGA BRCA data set."
)
```

>>>>

```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
if ( !file.exists( './data/exampleGenes.rds' ) )  {
    set.seed(01122021)
    exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
    saveRDS(exampleGenes, file = './data/exampleGenes.rds')
}
exampleGenes <- readRDS('./data/exampleGenes.rds')
## Add row names per WWP comment
small <-  brca %>% select( exampleGenes ) %>% as.data.frame
numPatients <- 5
smallForTable <- head( small, n = numPatients )
smallForTable$PatientID <- head( readLines('./data/examplePatients.txt'), n = numPatients )
smallForTable <- smallForTable[ , c(4, 1:3 ) ]
smallForTable[ , 2:4 ] <- round( smallForTable[ , 2:4 ] )
knitr::kable(smallForTable ,
      booktabs = TRUE,
      caption = "mRNA expression for three selected high-expressing genes (*RPL5*, *TKNIP*, and *VIM*) for the first five patients in the TCGA BRCA data set."
)
```

### 2021-01-13 020-simulation-algorithm.Rmd
2nd draft. Polishing ch050 after new RNA-seq

1. Mapping step  
   i. Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$.   
   or  
i$^\prime$. Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
2. Checking admissibility and nearest correlation computation
    (i) Check that $R_X$ is a correlation matrix, a PSD matrix with 1's along the diagonal.   
	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme.  
	Otherwise,  
(ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $R_X$, in the Frobenius norm.
3. NORTA step
	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
	(iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

>>>>

1. Mapping step  
   i. Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$.   
   or  
i$^\prime$. Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
2. Checking admissibility and, if needed, nearest correlation computation
    (i) Check that $R_X$ is a correlation matrix, a PSD matrix with 1's along the diagonal.   
	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme.  
	Otherwise,  
(ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $R_X$, in the Frobenius norm.
3. NORTA step
	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
	(iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

### 2021-01-13 050-examples.Rmd
2nd draft. Polishing ch050 after new RNA-seq

```{r ch050-readBRCA, echo=FALSE, eval=TRUE, cache=F}
## full workflow simulating RNA-seq data
brcaRDS <- 'data/brca95.rds'
myProb = 0.95
brca <- readRDS(file = brcaRDS)
d <- ncol(brca)
## smaller set for prototyping
## d=5
## brca <- brca[ , 1:d]
```

@@

```{r ch050-estMargins, echo = TRUE, eval = TRUE, cache=F}
sizes <- apply( unname(as.matrix(brca)), 2, nbinom_mom )[1, ]
probs <- apply( unname(as.matrix(brca)), 2, nbinom_mom )[2, ]
```

@@

```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache = FALSE}
## Set the number of random vectors (B in text)
n <- 10000
## construct margins
nb_margins <- make_nbinom_alist(sizes, probs)
## run sims
sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
                   ensure_PSD = TRUE, cores = cores) 
colnames(sim_nbinom) <- names(brca)
```

```{r ch050-runBRCA, echo=FALSE, results=FALSE, cache = FALSE}
nb_margins <- make_nbinom_alist(sizes, probs)
brcaSimRDS <- paste0( "results/brca", round(myProb*100), "sim.rds" )
if( !file.exists(brcaSimRDS) ) {
    ## Set the number of random vectors
    n <- 10000
    ## construct margins
    ## run sims
    set.seed(44)
    system.time( sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
                                    ensure_PSD = TRUE, cores = cores) )
    colnames(sim_nbinom) <- names(brca)
    saveRDS( sim_nbinom, brcaSimRDS )
}
sim_nbinom <- readRDS( brcaSimRDS )
```

@@

```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.'}
knitr::include_graphics('fig/ch050-simDataFig.png')
```

@@

```{r ch050-makeFigBRCA, echo=F, eval=T, cache=F}
if( !file.exists("fig/ch050-figBRCA.png") ) {
    ## estimate rho from sims and plot
    nb_Rho_hat <- cor_fast(sim_nbinom, method = corType)
    ## triplot
    ## png("fig/ch050-figBRCA.png", width = 4, res = 100, units = 'in')
    png("fig/ch050-figBRCA.png", res = 300, width = 5, height = 4, units = 'in')
    par(mfrow=c(1,3))
    plot(nb_Rho[lower.tri(nb_Rho)], nb_Rho_hat[lower.tri(nb_Rho_hat)],
         xlim = c(-1, 1), ylim = c(-1, 1),
         pch = 1, cex = 1, col = rgb(0, 0, 0, 0.05),
         xlab = "Target Correlation", ylab = "Estimated Correlation")
         ## main = "Spearman Cor. Matching")
    abline(a = 0, b = 1, col="red", lty="dashed")
    ## estimate prob,size from sims and plot
    nbinom_size_prob_hat <- apply(sim_nbinom, 2, nbinom_mom)
    plot(sizes, nbinom_size_prob_hat[1,],
         pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
         xlab = "Target Size", ylab = "Estimated Size")
         ## main = "Marginal NB Size Matching")
    abline(a=0, b=1, col="red", lty="dashed")
    plot(log(probs), log(nbinom_size_prob_hat[2,]),
         pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
         xlab = "Target Log Probability", ylab = "Estimated Log Probability")
         ## main = "Marginal NB Prob Matching")
    abline(a=0, b=1, col="red", lty="dashed")
    dev.off()
    par(mfrow=c(1,1))
}
```

>>>>

```{r ch050-readBRCA, echo=FALSE, eval=TRUE, cache=F}
d <- 1000
brcaRDS <- paste0( "data/brca", d, ".rds" )
brca <- readRDS(file = brcaRDS)
## smaller set for prototyping
## d=5
## brca <- brca[ , 1:d]
```

@@

```{r ch050-estMargins, echo = TRUE, eval = TRUE, cache=F}
sizes <- apply( unname(as.matrix(brca)), 2, nbinom_mom )[1, ]
probs <- apply( unname(as.matrix(brca)), 2, nbinom_mom )[2, ]
## construct margins
nb_margins <- make_nbinom_alist(sizes, probs)
```

@@

Now we check admissibility of specified correlation matrix.
We begin with the pairwise, 2D MC estimated bounds.
And then follow the `r d`D bounds.

```{r ch050-cor-check, echo=TRUE, eval=TRUE, cache = FALSE}
## 1. Mapping step first
R_X <- cor_convert(rho = R_S, from = type, to = 'pearson')
## 2a. Check admissibility
is_valid_correlation <- function(rho) {
  all(diag(rho) == 1) && matrixcalc::is.positive.semi.definite(rho)
}
is_valid_correlation( R_X )

## 2b. compute nearest correlation
if ( ! is_valid_correlation( R_X ) ) {
    ## system.time( R_Xtilde <- cor_nearPSD( G = R_X )  )
    R_Xtilde <- cor_nearPSD( G = R_X )
    ## Quantify the error
    targets <- R_X[ lower.tri( R_X ) ]
    approximates <- R_Xtilde[ lower.tri( R_Xtilde ) ]
    ## R_Xtilde[1:5, 1:5]
    ## norm( x = R_X - R_Xtilde, type = "F" )
    R_X  <- R_Xtilde
}
## qplot( x = targets, y = approximates ) + geom_abline(slope=1,intercept = 0)
## qplot( x = targets - approximates ) + geom_abline(slope=1,intercept = 0)
summary(targets - approximates)
```

While the exact dxd Spearman correlation matrix is not strictly admissible in our scheme (seen by the non-PSD result above), the approximation is of sufficient quality for most applications with a maximum pairwise of `r max( targets - approximations)` and average error of `r mean( targets - approximations)`.

```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache = FALSE}
sim_nbinom <- rvec(10000, R_X, nb_margins, ensure_PSD = FALSE, cores = cores) 
colnames(sim_nbinom) <- names(brca)
```

```{r ch050-runBRCA, echo=FALSE, results=FALSE, cache = FALSE}
brcaSimRDS <- paste0( "results/brca", d, "sim.rds" )
if( !file.exists(brcaSimRDS) ) {
    ## Set the number of random vectors
    n <- 10000
    ## construct margins
    ## run sims
    set.seed(44)
    sim_nbinom <- rvec(10000, R_X, nb_margins, ensure_PSD = FALSE, cores = cores)
    saveRDS( sim_nbinom, brcaSimRDS )
}
sim_nbinom <- readRDS( brcaSimRDS )
```

@@

```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes generally replicates the estimated data structure. The data do not exhibit outlying points, but do possess the desired Spearman correlations, central tendencies, and discrete values.'}
knitr::include_graphics('fig/ch050-simDataFig.png')
```

@@

```{r ch050-makeFigBRCA, echo=F, eval=T, cache=F}
if( !file.exists("fig/ch050-figBRCA.png") ) {
    ## estimate rho from sims and plot
    R_S_hat <- cor_fast(sim_nbinom, method = type)
    ## triplot
    ## png("fig/ch050-figBRCA.png", width = 4, res = 100, units = 'in')
    png("fig/ch050-figBRCA.png", res = 300, width = 5, height = 4, units = 'in')
    par(mfrow=c(1,3))
    plot(R_S[lower.tri(R_S)], R_S_hat[lower.tri(R_S_hat)],
         xlim = c(-1, 1), ylim = c(-1, 1),
         pch = 1, cex = 1, col = rgb(0, 0, 0, 0.05),
         xlab = "Target Correlation", ylab = "Estimated Correlation")
         ## main = "Spearman Cor. Matching")
    abline(a = 0, b = 1, col="red", lty="dashed")
    ## estimate prob,size from sims and plot
    nbinom_size_prob_hat <- apply(sim_nbinom, 2, nbinom_mom)
    plot(sizes, nbinom_size_prob_hat[1,],
         pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
         xlab = "Target Size", ylab = "Estimated Size")
         ## main = "Marginal NB Size Matching")
    abline(a=0, b=1, col="red", lty="dashed")
    plot(log(probs), log(nbinom_size_prob_hat[2,]),
         pch = 1, cex = 1, col = rgb(0, 0, 0, 1),
         xlab = "Target Log Probability", ylab = "Estimated Log Probability")
         ## main = "Marginal NB Prob Matching")
    abline(a=0, b=1, col="red", lty="dashed")
    dev.off()
    par(mfrow=c(1,1))
}
```
